---
title: "Dplyr `select()`"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Use summary functions to assess the structure of a data frame.  
2. Us the select function of `dplyr` to build data frames restricted to variable of interest.  
3. Use the `rename()` function to provide new, consistent names to variables in data frames.  

## Load the tidyverse
For the remainder of the quarter, we will work within the `tidyverse`. At the start of each lab, the library needs to be loaded as shown below.  
```{r message=FALSE, warning=FALSE}
library("tidyverse")
```

## Load the data
These data are from: Gaeta J., G. Sass, S. Carpenter. 2012. Biocomplexity at North Temperate Lakes LTER: Coordinated Field Studies: Large Mouth Bass Growth 2006. Environmental Data Initiative.  [link](https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-ntl&identifier=267)  
```{r}
fish <- read_csv("data/Gaeta_etal_CLC_data.csv")
```

## Data Structure
Once data have been uploaded, let's get an idea of its structure, contents, and dimensions.  
```{r}
glimpse(fish)
```

```{r}
summary(fish) # math can be a little strange, be careful with math 
```

## Tidyverse
The [tidyverse](www.tidyverse.org) is a collection of packages that make workflow in R easier. The packages operate more intuitively than base R commands and share a common organizational philosophy.  

base r code has its own structure and organization compared to tidyverse


## dplyr
The first package that we will use that is part of the tidyverse is `dplyr`. `dplyr` is used to transform data frames by extracting, rearranging, and summarizing data such that they are focused on a question of interest. This is very helpful,  especially when wrangling large data, and makes dplyr one of most frequently used packages in the tidyverse. The two functions we will use most are `select()` and `filter()`.  

## `select()`
Select allows you to pull out columns of interest from a dataframe. To do this, just add the names of the columns to the `select()` command. The order in which you add them, will determine the order in which they appear in the output.

Pulls out variables of interest

```{r}
names(fish)
```

We are only interested in lakeid and scalelength.
```{r}
select(fish, "lakeid", "scalelength") # first put the data you want to search "fish"
```

Remember that you can save the output to a new dataframe.
```{r}
fish_subset <- select(fish, "lakeid", "scalelength")
```

To add a range of columns use `start_col:end_col`.
```{r}
select(fish, fish_id:length) # for working on numerically sequences you want to pull out 
```

The - operator is useful in select. It allows us to select everything except the specified variables.
```{r}
select(fish, -"fish_id", -"annnumber", -"length", -"radii_length_mm") # pull out all the variables (-) except, working with big data frames
```

For very large data frames with lots of variables, `select()` utilizes lots of different operators to make things easier. Let's say we are only interested in the variables that deal with length.
```{r}
select(fish, contains("length")) # give me tha names that contain the strings "length" for variable
```

When columns are sequentially named, `starts_with()` makes selecting columns easier.
```{r}
select(fish, starts_with("radii")) # will pull out variables that stards with "radii"
```

Options to select columns based on a specific criteria include:  
1. ends_with() = Select columns that end with a character string  
2. contains() = Select columns that contain a character string  
3. matches() = Select columns that match a regular expression  
4. one_of() = Select columns names that are from a group of names  

We can use these functions to select columns that end with "id".
```{r}
select(fish, ends_with("id"))
```

Or, we can select columns that contain "fish".
```{r}
select(fish, contains("fish"))
```

You can also select columns based on the class of data.  
```{r}
glimpse(fish)
```

```{r}
select_if(fish, is.numeric) # pulls out variables that are numeric 
```

To select all columns that are *not* a class of data, you need to add a `~`.
```{r}
select_if(fish, ~!is.numeric(.)) #the ! means not , search all variables and check in each one thats not numeric 
```

## Practice  
For this exercise we will use life history data `mammal_lifehistories_v2.csv` for mammals. The [data](http://esapubs.org/archive/ecol/E084/093/) are from:  
**S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.**  

Load the data.  
```{r}
mammals <- read_csv("data/mammal_lifehistories_v2.csv")
```

1. Use one or more of your favorite functions to assess the structure of the data.  
```{r}
glimpse(mammals)
```
```{r}
summary(mammals)
```

2. Are there any NAs? Are you sure? Try taking an average of `max. life` as a test.  
```{r}
anyNA(mammals)
mean(mammals, max.life)
```

3. What are the names of the columns in the `mammals` data?
```{r}
names(mammals)
```

4. Rename any columns that have capitol letters or punctuation issues.  
```{r}
mammals_updated <- rename(mammals, genus = 'Genus', litter_size = "litter size", wean_mass = 'wean mass', litters_per_year = 'litters/year', afr = 'AFR', max_life = "max. life") # we only run and rename once

summary(mammals_updated)
```

5. We are only interested in the variables `genus`, `species`, and `mass`. Use `select()` to build a new dataframe `mass` focused on these variables.
```{r}
mass <- select(mammals_updated, genus, species, mass_life)
mass # prints the df, not something you want to do in your hw 
```

6. What if we only wanted to exclude `order` and `family`? Use the `-` operator to make the code efficient.
```{r}
select(mammals_updated, -"order", -"family")
```

7. Select the columns that include "mass" as part of the name.  
```{r}
select(mammals_updated, includes("mass") )
```

8. Select all of the columns that are of class `character`.  
```{r}
select(mammals_updated, ~!is.numeric(.))
```

## Janitor
Another option is to use the janitor package. The janitor package has a function called `clean_names()` that will make all column names lowercase and replace spaces with underscores.
```{r}
#install.packages("janitor")
library(janitor)
```

```{r}
names(mammals)
```

```{r}
clean_names(mammals) # keeps data consistent making sure there are no inconsitencies with the name 
```

## That's it! Let's take a break and then move on to part 2! 

-->[Home](https://jmledford3115.github.io/datascibiol/)  